Forcing for Isabelle

### 2018-02-12 blueprint.tex
typo
first insight on why this is so damn difficult

Most probably, we'll need two or three versions of each of these
definitions: One internalized as an element of \formula;  a relativized
one with a class parameter; and finally one without
relativization. This is done in order to show absolutness and to take
advantage of the Isabelle/ZF ability to operate with terms of type
$\tyi$.

@@

Under the assumption of transitivity of $M$, the three versions are
equivalent. This is an \emph{absolutness result}:
\[
\phi(p,r) \lsii \phi^M(p,r) \lsii M;[p,r]\models\quine{\phi},
\]
when $p,r\in M$, and the list $[p,r]$ is the environment or
interpretation of free variables.

@@

\section{Names}
\begin{framed}
  The following is mostly independent of the previous section, but 
  \begin{bfseries}
    we assume that $\1\in\PP\in M$, $M$ is transitive and $1\in
    G\sbq\PP$.
  \end{bfseries}
  (Note that, in general, $G\notin M$.)
\end{framed}

@@

The definition of $\val$ proceeds by well founded recursion on
\verb|eee|$(M)$. Determining  $\val(\tau,\PP,G)$ uses the values
$\val(\sigma,\PP,G)$ for smaller $\sigma$:
\[
\sigma\in\{\sigma\}\in\lb\sigma,p\rb\in\tau.
\]

@@

We will now make more precise the definition of the map
$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
$\phi'$ will have $n+4$ free variables, where the first four account
for the forcing notion and a particular condition. If
$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
\[
p\forces_{\PP,\leq,\1}^* \phi(\tau_1,\dots,\tau_n).
\]
Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. For the
time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
variables. Informally, we call $\forces$ the \emph{forcing relation}.
Given $\quine\phi\in\formula$, we will write the forcing relation in
Isabelle/ZF by using  
\[
\forceisa(\quine\phi,\PP,\leq,\1)
\]

The base case (of atomic formulas) is famously the most
complicated one.

>>>>

Most probably, we'll need two or three versions of each of these
definitions: One internalized as an element of \formula;  a relativized
one with a class parameter; and finally one without
relativization. This is done in order to show absoluteness and to take
advantage of the Isabelle/ZF ability to operate with terms of type
$\tyi$.

@@

Under the assumption of transitivity of $M$, the three versions are
equivalent. This is an \emph{absoluteness result}:
\[
\phi(p,r) \lsii \phi^M(p,r) \lsii M;[p,r]\models\quine{\phi},
\]
when $p,r\in M$, and the list $[p,r]$ is the environment or
interpretation of free variables.

@@

\section{Names}
\begin{framed}
  The following is mostly independent of the previous section, but %
  \begin{bfseries}%
    we assume that $\1\in\PP\in M$, $M$ is transitive and 
    $1\in G\sbq\PP$.
  \end{bfseries}
  (Note that, in general, $G\notin M$.)
\end{framed}

@@

The definition of $\val$ proceeds by well-founded recursion on
\verb|eee|$(M)$. Determining  $\val(\tau,\PP,G)$ uses the values
$\val(\sigma,\PP,G)$ for smaller $\sigma$:
\[
\sigma\in\{\sigma\}\in\lb\sigma,p\rb\in\tau.
\]

@@

We will now make more precise the definition of the map
$\phi\mapsto\phi'$ and how it relates satisfaction in $M$ to that in
$M[G]$. Actually, if the formula $\phi$ has $n$ free variables,
$\phi'$ will have $n+4$ free variables, where the first four account
for the forcing notion and a particular condition. If
$\phi=\phi(x_1,\dots,x_n)$, the standard notation for $\phi'$ is
\[
p\forces_{\PP,\leq,\1}^* \phi(t_1,\dots,t_n).
\]
Here, $\PP,\leq,\1,$ and $p\in\PP$ are the extra parameters. 
%% For the
%% time being, we just consider $\tau_1,\dots,\tau_n$ to be ordinary
%% variables. 
Informally, we call $\forces$ the \emph{forcing relation}.
Given $\quine\phi\in\formula$, we will write the forcing relation in
Isabelle/ZF by using  
\[
\forceisa(\quine\phi,\PP,\leq,\1,p)
\]

The base case (for atomic formulas) is famously the most
complicated one. 
\begin{multline}
  \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q)\lsii 
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q),
\end{multline}
\begin{multline}
  \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  \forceisa(\quine{t_1 =s},\PP,\leq,\1,q))\} \\
  \text{is dense below $p$.''}
\end{multline}

Actually, these cases depend on the theorem of well-founded recursion,
so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
must be obtained from the proof of that theorem. This means that we need
an internalized version of the theorem.
